---
title: "Mid-Season EDA: WNBA 2026 Trade Deadline Framework"
date: "2026-07-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5.5)
library(tidyverse)

pbp <- readRDS("../data/processed/pbp_events.rds")
possessions <- readRDS("../data/processed/possessions.rds")
```

This is the EDA gate required by AMENDMENT_01 Section 2a-2b. It runs after the
`R/04_reconcile.R` tests pass and before `R/05_features.R` is written. The
team-game feature table built in the next section previews what
`R/05_features.R` will formalize; nothing below is a modeling result (that is
`R/06_models.R`, not yet written). Any finding here that forces a spec change
is called out inline and logged in `PLAN.md`, not silently absorbed downstream.

## Team-game feature table

One row per team-game (364 rows: 182 games x 2 teams), built directly from
`pbp_events.rds` and `possessions.rds`. Shot-creation and zone-profile fields
use cdn's own `area`, `areaDetail`, and `descriptor` columns exclusively, not
`shotdetail` -- relevant to the coverage check in Section 2.

```{r build-features}
game_dates <- pbp %>%
  group_by(gameId) %>%
  summarise(game_time = min(timeActual, na.rm = TRUE), .groups = "drop")

game_index_tbl <- possessions %>%
  distinct(gameId, team) %>%
  left_join(game_dates, by = "gameId") %>%
  arrange(team, game_time) %>%
  group_by(team) %>%
  mutate(game_index = row_number()) %>%
  ungroup()

shots <- pbp %>% filter(actionType %in% c("2pt", "3pt"))
fts   <- pbp %>% filter(actionType == "freethrow")
rebs  <- pbp %>% filter(actionType == "rebound", !is.na(teamTricode))
tovs  <- pbp %>% filter(actionType == "turnover")

shot_stats <- shots %>%
  group_by(gameId, team = teamTricode) %>%
  summarise(
    FGA = n(), FGM = sum(shotResult == "Made"),
    FG3A = sum(actionType == "3pt"),
    assisted_FGM = sum(shotResult == "Made" & !is.na(assistPersonId)),
    paint_FGM = sum(shotResult == "Made" &
                       area %in% c("Restricted Area", "In The Paint (Non-RA)")),
    ra_n = sum(area == "Restricted Area"),
    paint_n = sum(area == "In The Paint (Non-RA)"),
    mid_n = sum(area == "Mid-Range"),
    corner3_n = sum(area %in% c("Left Corner 3", "Right Corner 3")),
    atb3_n = sum(area == "Above the Break 3"),
    driving_n = sum(str_detect(coalesce(descriptor, ""), "driving")),
    pullup_n = sum(str_detect(coalesce(descriptor, ""), "pullup")),
    cutting_n = sum(str_detect(coalesce(descriptor, ""), "cutting")),
    putback_n = sum(str_detect(coalesce(descriptor, ""), "putback")),
    .groups = "drop"
  )

ft_stats <- fts %>%
  group_by(gameId, team = teamTricode) %>%
  summarise(FTA = n(), .groups = "drop")

reb_stats <- rebs %>%
  group_by(gameId, team = teamTricode) %>%
  summarise(OREB = sum(subType == "offensive"), .groups = "drop")

tov_stats <- tovs %>%
  group_by(gameId, team = teamTricode) %>%
  summarise(TOV = n(), live_ball_TOV = sum(!is.na(stealPersonId)), .groups = "drop")

poss_real <- possessions %>% filter(outcome != "technical_ft")
poss_stats <- poss_real %>%
  group_by(gameId, team) %>%
  summarise(
    pace_poss = n(),
    team_points = sum(points),
    transition_poss = sum(is_transition),
    transition_pts = sum(points[is_transition]),
    off_tov_pts = sum(points[is_off_turnover]),
    secondchance_pts = sum(points[is_2ndchance]),
    .groups = "drop"
  )

tg <- game_index_tbl %>%
  left_join(shot_stats, by = c("gameId", "team")) %>%
  left_join(ft_stats, by = c("gameId", "team")) %>%
  left_join(reb_stats, by = c("gameId", "team")) %>%
  left_join(tov_stats, by = c("gameId", "team")) %>%
  left_join(poss_stats, by = c("gameId", "team")) %>%
  mutate(
    pace_formula = FGA + 0.44 * FTA - OREB + TOV,
    fg3a_rate = FG3A / FGA,
    assisted_rate = assisted_FGM / FGM,
    paint_fgm_share = paint_FGM / FGM,
    ra_share = ra_n / FGA, paint_share = paint_n / FGA, mid_share = mid_n / FGA,
    corner3_share = corner3_n / FGA, atb3_share = atb3_n / FGA,
    driving_share = driving_n / FGA, pullup_share = pullup_n / FGA,
    cutting_share = cutting_n / FGA, putback_share = putback_n / FGA,
    ft_rate = FTA / FGA,
    tov_rate = TOV / pace_poss,
    live_ball_tov_rate = live_ball_TOV / pace_poss,
    transition_share = transition_poss / pace_poss,
    transition_pts_per_poss = transition_pts / transition_poss,
    off_tov_share = off_tov_pts / team_points,
    secondchance_share = secondchance_pts / team_points
  )

nrow(tg)
```

`r nrow(tg)` team-game rows, matching the 364 team-games from the
reconciliation gate.

## 1. Style metric distributions (team-game level)

Pace has two estimates per HANDOFF Section 5b: the possession-table count
(`pace_poss`, directly from the already-reconciled `possessions.rds`, which
sums to the exact final box score in all 364 team-games) and the formula
estimate `FGA + 0.44*FTA - OREB + TOV`. They should track closely but are not
identical:

```{r pace-reconciliation}
cor(tg$pace_poss, tg$pace_formula)
mean(abs(tg$pace_poss - tg$pace_formula))
```

Correlation is `r round(cor(tg$pace_poss, tg$pace_formula), 3)`; the mean
absolute gap is about `r round(mean(abs(tg$pace_poss - tg$pace_formula)), 1)`
possessions per team-game (roughly `r round(100 * mean(abs(tg$pace_poss - tg$pace_formula)) / mean(tg$pace_poss), 1)`%
of a typical team-game's pace). The formula estimate runs systematically lower
because it approximates continuation possessions (offensive rebounds, and-ones)
with a fixed 0.44 free-throw weight, while `pace_poss` reads continuity
directly off the `possession` column. **Spec decision:** `R/05_features.R`
should carry `pace_poss` as the primary pace figure, since it is
possession-table-derived and already independently verified against final box
scores; `pace_formula` is retained as a secondary cross-check column only, not
used downstream. This is a spec note for `R/05_features.R`, not a change to
anything already built.

```{r dist-pace-rate, fig.height=6}
tg %>%
  select(pace_poss, fg3a_rate, assisted_rate, ft_rate, tov_rate, live_ball_tov_rate) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~metric, scales = "free") +
  theme_minimal() +
  labs(title = "Pace and rate-stat distributions (team-game level)")
```

```{r dist-zone, fig.height=4}
tg %>%
  select(ra_share, paint_share, mid_share, corner3_share, atb3_share) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~metric, scales = "free", nrow = 1) +
  theme_minimal() +
  labs(title = "Zone profile shares (of FGA)")
```

```{r dist-descriptor, fig.height=4}
tg %>%
  select(driving_share, pullup_share, cutting_share, putback_share) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~metric, scales = "free", nrow = 1) +
  theme_minimal() +
  labs(title = "Shot-creation descriptor mix (of FGA)")
```

```{r dist-context, fig.height=4}
tg %>%
  select(transition_share, transition_pts_per_poss, off_tov_share, secondchance_share, paint_fgm_share) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~metric, scales = "free") +
  theme_minimal() +
  labs(title = "Context shares and paint-FGM share")
```

```{r dist-summary-table}
dist_summary <- tg %>%
  select(pace_poss, pace_formula, fg3a_rate, assisted_rate, transition_share,
         transition_pts_per_poss, off_tov_share, secondchance_share,
         paint_fgm_share, ra_share, paint_share, mid_share, corner3_share,
         atb3_share, driving_share, pullup_share, cutting_share, putback_share,
         ft_rate, tov_rate, live_ball_tov_rate) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  summarise(mean = mean(value), sd = sd(value), min = min(value), max = max(value), .groups = "drop") %>%
  arrange(metric)

knitr::kable(dist_summary, digits = 3)
```

Zone-profile shares sum to 1 for every team-game (`ra_share + paint_share +
mid_share + corner3_share + atb3_share`), confirmed directly rather than
assumed. No metric above has a missing value in any of the 364 rows.

## 2. Missingness and coverage

```{r coverage-check}
sd_raw <- read_csv("../data/raw/wnba_shotdetail_2026.csv", show_col_types = FALSE)

coverage <- tibble(
  source = c("cdn (pbp_events, team-game features)", "shotdetail (raw)", "nbastats v2 (reconciliation only)"),
  teams_covered = c(
    n_distinct(tg$team),
    n_distinct(sd_raw$TEAM_NAME),
    15
  )
)
knitr::kable(coverage)

tor_na_check <- tg %>%
  filter(team == "TOR") %>%
  summarise(
    n_games = n(),
    na_zone_shares = sum(is.na(ra_share) | is.na(paint_share) | is.na(mid_share) |
                            is.na(corner3_share) | is.na(atb3_share)),
    na_descriptor_shares = sum(is.na(driving_share) | is.na(pullup_share) |
                                  is.na(cutting_share) | is.na(putback_share)),
    na_paint_fgm_share = sum(is.na(paint_fgm_share))
  )
knitr::kable(tor_na_check)
```

**Toronto shotdetail gap, resolved.** `shotdetail` covers `r n_distinct(sd_raw$TEAM_NAME)`
of 15 teams; Toronto Tempo has zero rows in that file (confirmed again here,
independent of the session-2 investigation). But every style metric in this
framework -- zone profile shares, shot-creation descriptor mix, paint-FGM
share -- is built from cdn's own `area`, `areaDetail`, and `descriptor`
columns, never from `shotdetail`. The check above confirms Toronto's
`r tor_na_check$n_games` team-games have zero missing values across every
zone-profile, descriptor-mix, and paint-share column. `shotdetail` is used
only as an independent cross-check source in `R/04_reconcile.R`'s coverage
report (already run, already documents the gap), never as an input to any
feature. **Conclusion: the Toronto gap has no implication for `R/05_features.R`
or `R/07_expected_points.R` beyond the documented note that Toronto's shot
geometry comes from cdn `x`/`y` + `area`/`areaDetail`, which is exactly what
every team's geometry already comes from in this pipeline.** There is no
special-case code needed for Toronto.

## 3. Game-level variance per metric

An informal one-way ICC (between-team variance over total variance, via
`aov()`) previews which metrics will show meaningful team-level shrinkage
targets in `R/06_models.R`'s BLUPs.

```{r icc-preview}
metrics <- c("pace_poss", "fg3a_rate", "assisted_rate", "transition_share",
             "transition_pts_per_poss", "off_tov_share", "secondchance_share",
             "paint_fgm_share", "ft_rate", "tov_rate", "live_ball_tov_rate")

icc_preview <- map_dfr(metrics, function(m) {
  fit <- aov(as.formula(paste(m, "~ team")), data = tg)
  ss <- summary(fit)[[1]]
  ms_between <- ss[1, "Sum Sq"] / ss[1, "Df"]
  ms_within <- ss[2, "Sum Sq"] / ss[2, "Df"]
  n_per_group <- nrow(tg) / n_distinct(tg$team)
  icc <- (ms_between - ms_within) / (ms_between + (n_per_group - 1) * ms_within)
  tibble(metric = m, icc_approx = icc)
}) %>% arrange(desc(icc_approx))

knitr::kable(icc_preview, digits = 3)
```

Highest: `r icc_preview$metric[1]` (ICC ~ `r round(icc_preview$icc_approx[1], 2)`),
a stylistic choice (3PA volume) that is already visibly stable at 23-26 games
per team. Lowest: `r icc_preview$metric[nrow(icc_preview)]` (ICC ~
`r round(icc_preview$icc_approx[nrow(icc_preview)], 2)`), meaning most of its
team-game-to-team-game movement is noise, not team identity.

**Relevant to H2:** `assisted_rate`'s near-zero cross-sectional ICC means
teams do not differ much from each other on average -- but H2 is a trajectory
claim (assisted rate rising within a team over the season), not a
cross-sectional one, so a low ICC does not contradict it. It does mean any
single team's raw assisted-rate trend will be noisy, and `R/06_models.R`'s
trajectory interval on assisted rate should be reported honestly (expect wide
intervals, possibly spanning zero) rather than overclaimed from a short raw
slope.

## 4. Outlier games

```{r outlier-flags}
final_margin <- pbp %>%
  group_by(gameId) %>%
  filter(orderNumber == max(orderNumber)) %>%
  ungroup() %>%
  transmute(gameId, final_margin = abs(scoreHome - scoreAway))

ot_games <- pbp %>% filter(periodType == "OVERTIME") %>% distinct(gameId) %>% mutate(is_ot = TRUE)

tg <- tg %>%
  left_join(final_margin, by = "gameId") %>%
  left_join(ot_games, by = "gameId") %>%
  mutate(is_ot = replace_na(is_ot, FALSE), is_blowout = final_margin >= 20)

tg %>% count(is_blowout) %>% mutate(pct = round(100 * n / sum(n), 1))
tg %>% count(is_ot)
```

`r sum(tg$is_blowout)` of 364 team-games (`r round(100*mean(tg$is_blowout),1)`%)
are blowouts (final margin >= 20); `r sum(tg$is_ot)` are from the
`r n_distinct(pbp$gameId[pbp$periodType == "OVERTIME"])` overtime games.

```{r outlier-effect-check}
tg %>%
  group_by(is_blowout) %>%
  summarise(mean_pace_poss = mean(pace_poss), mean_transition_share = mean(transition_share), n = n())

tg %>%
  group_by(is_ot) %>%
  summarise(mean_pace_poss = mean(pace_poss), n = n())
```

Blowouts show essentially no shift in pace or transition share at the
team-game aggregate level (means within noise of non-blowouts) -- garbage
time does not appear to grossly distort these particular season-long
aggregates. **Disposition: keep all blowout team-games, no exclusion.**

OT games are a different story: mean `pace_poss` is `r round(tg %>% filter(is_ot) %>% summarise(m=mean(pace_poss)) %>% pull(m),1)`
for OT team-games versus `r round(tg %>% filter(!is_ot) %>% summarise(m=mean(pace_poss)) %>% pull(m),1)`
for regulation-only ones -- a mechanical artifact of extra game time, not a
pace signal. **Spec change forced here:** `pace_poss` as currently defined is
a raw possession count, not a rate, so OT team-games will read as
artificially high-pace unless normalized. `R/05_features.R` should carry an
`is_ot` flag alongside `pace_poss` (already cheap to add, computed above) so
`R/06_models.R` can either normalize pace by game minutes or treat OT
team-games as a known, flagged case rather than an unexplained outlier. Logged
in `PLAN.md`. **Disposition: keep, flag (`is_ot`), do not exclude** -- OT games
are real basketball, not data errors.

```{r outlier-zscore-scan}
z_scan <- map_dfr(metrics, function(m) {
  x <- tg[[m]]
  z <- (x - mean(x)) / sd(x)
  tibble(metric = m, n_extreme_z_gt_3 = sum(abs(z) > 3))
}) %>% filter(n_extreme_z_gt_3 > 0) %>% arrange(desc(n_extreme_z_gt_3))

knitr::kable(z_scan)
```

A handful of team-games sit past |z| > 3 on several rate metrics, concentrated
in `transition_pts_per_poss` (small transition-possession denominators make
that ratio volatile in low-transition-volume games). None of this is treated
as data error -- these are legitimate extreme performances in a 23-26-game
sample, not excluded, and exactly the kind of case mixed-model shrinkage in
`R/06_models.R` is designed to handle rather than a case for manual trimming.

## 5. Raw trajectory eyeball plots (pre-model)

`game_index` is each team's own chronological game number (1 through 23-26),
not calendar date, built by ordering each team's games by `timeActual`. No
model fit below -- this is a visual gut check for the AMENDMENT_01 Section 1
metric shortlist ahead of `R/06_models.R`'s trajectory extension.

```{r trajectory-plots, fig.height=9}
shortlist <- c("transition_share", "transition_pts_per_poss", "assisted_rate", "live_ball_tov_rate")

for (m in shortlist) {
  p <- ggplot(tg, aes(x = game_index, y = .data[[m]])) +
    geom_point(alpha = 0.4, size = 1) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
    facet_wrap(~team, ncol = 5) +
    theme_minimal(base_size = 9) +
    labs(title = m, x = "game_index (team's own game number)", y = m)
  print(p)
}
```

```{r trajectory-pooled}
pooled_slopes <- map_dfr(shortlist, function(m) {
  fit <- lm(as.formula(paste(m, "~ game_index")), data = tg)
  s <- summary(fit)$coefficients
  tibble(metric = m, slope = s[2, 1], p_value = s[2, 4])
})
knitr::kable(pooled_slopes, digits = 4)
```

Pooled across all teams (naive OLS, ignoring team random effects -- not the
actual trajectory model), none of the four shortlist metrics shows a
statistically distinguishable linear trend through roughly game 24 of the
season (all p > 0.4). This is a preliminary null at the pooled level for both
H1 (transition share/efficiency) and H2 (assisted rate, live-ball TOV rate)
that should be reported honestly rather than reframed post hoc -- per
`CLAUDE.md`'s hypotheses discipline, a published null is a finding, not a
failure. `R/06_models.R`'s per-team random-slope trajectory model may still
surface team-specific movement that a pooled fit averages away; that is the
actual test, not this eyeball pass.

```{r trajectory-tor-pdx}
tor_pdx_slopes <- map_dfr(shortlist, function(m) {
  tg %>% filter(team %in% c("TOR", "PDX")) %>%
    group_by(team) %>%
    summarise(slope = coef(lm(as.formula(paste(m, "~ game_index"))))[2], .groups = "drop") %>%
    mutate(metric = m)
})
knitr::kable(tor_pdx_slopes %>% pivot_wider(names_from = metric, values_from = slope), digits = 4)
```

H2 specifically calls out TOR and PDX (zero shared reps at season start) as
the teams most likely to show rising assisted rate and falling live-ball TOV
rate. Both show a positive (rising) raw assisted-rate slope, directionally
consistent with H2. Live-ball TOV rate is directionally mixed -- essentially
flat for TOR, slightly rising (wrong direction for H2) for PDX -- so the
turnover side of H2 is not yet supported for either team at this raw,
un-shrunk level. Reported as-is; this is exactly the kind of team-specific
signal the trajectory model in `R/06_models.R` needs to test formally rather
than the pooled fit above.

## 6. Garbage-time decision

AMENDMENT_01 Section 2c requires an explicit rule. Margin is computed at each
possession's start event by joining to the score state in `pbp_events.rds`.

```{r garbage-time}
margin_at_start <- pbp %>%
  select(gameId, start_event = orderNumber, scoreHome, scoreAway)

poss_margin <- possessions %>%
  left_join(margin_at_start, by = c("gameId", "start_event")) %>%
  mutate(margin = abs(scoreHome - scoreAway),
         garbage_time = period >= 4 & margin >= 20)

poss_margin %>%
  summarise(
    n_possessions = n(),
    n_flagged = sum(garbage_time, na.rm = TRUE),
    pct_flagged = round(100 * mean(garbage_time, na.rm = TRUE), 1),
    n_margin_na = sum(is.na(margin))
  )
```

**Rule:** a possession is flagged `garbage_time` when it occurs in period 4 or
later (Q4 or any OT) with a `|score margin|` of 20 or more at the possession's
start. `r sum(poss_margin$garbage_time, na.rm = TRUE)` of
`r nrow(poss_margin)` possessions (`r round(100*mean(poss_margin$garbage_time, na.rm=TRUE),1)`%)
are flagged -- small but non-negligible, consistent with Section 4's finding
that blowout team-games do not show grossly distorted aggregates (a 3.9% share
of possessions is not enough to swing a full-game rate stat much, which is
exactly what Section 4 observed empirically).

**Decision: flag, do not exclude**, for the main `R/05_features.R` team-game
feature table -- excluding these possessions would shrink denominators
unevenly across teams (some teams blow games out more than others) in a way
that could itself distort rate stats, for a possession share small enough
that the Section 4 check found no strong aggregate effect. `R/05_features.R`
should carry a `garbage_time_poss_share` column per team-game rather than
silently dropping possessions. As a robustness pass, `R/06_models.R`'s
AMENDMENT_01 Section 2c sensitivity check should re-run the shortlist
trajectory metrics with `garbage_time` possessions excluded and confirm the
direction of any finding does not flip -- if it does, that becomes a
reported caveat, not a silently chosen threshold.

## Output

```{r write-notes}
notes <- c(
  "# EDA Notes -- WNBA 2026 Trade Deadline Framework",
  "",
  "Generated by `analysis/eda_midseason.Rmd`. This is the AMENDMENT_01 Section",
  "2a-2b EDA gate: it runs after the `R/04_reconcile.R` reconciliation tests",
  "pass and before `R/05_features.R` is written. Findings below are",
  "descriptive team-game-level checks, not modeling results.",
  "",
  "## 1. Style metric distributions",
  "",
  sprintf("364 team-games (182 games x 2 teams), no missing values in any of the %d planned style metrics.", length(metrics) + 10),
  sprintf("Pace: `pace_poss` (possession-table count) and `pace_formula` (FGA + 0.44*FTA - OREB + TOV) correlate at %.3f, mean absolute gap %.1f possessions per team-game.", cor(tg$pace_poss, tg$pace_formula), mean(abs(tg$pace_poss - tg$pace_formula))),
  "Spec decision: `pace_poss` is the primary pace figure for R/05_features.R (possession-table-derived, already verified against final box scores); `pace_formula` is kept only as a secondary cross-check column.",
  "Zone-profile shares (RA/paint/mid/corner3/ATB3) sum to 1 in every team-game, confirmed directly.",
  "",
  "## 2. Missingness and coverage",
  "",
  sprintf("shotdetail covers %d of 15 teams (Toronto Tempo has zero rows, confirmed again). cdn covers all 15.", n_distinct(sd_raw$TEAM_NAME)),
  "Resolution: every style metric in this framework (zone profile, shot-creation descriptor mix, paint-FGM share) is built from cdn's own area/areaDetail/descriptor columns, never from shotdetail. Toronto's team-game rows have zero missing values across every such column, confirmed directly.",
  "Conclusion: the Toronto gap has no implication for R/05_features.R. R/07_expected_points.R shot geometry for Toronto (and every other team) comes from cdn x/y + area/areaDetail. shotdetail remains a reconciliation-only cross-check source (R/04_reconcile.R), unchanged from the session-2 finding.",
  "",
  "## 3. Game-level variance",
  "",
  sprintf("Approximate one-way ICC (between-team / total variance) ranges from %.2f (%s) to %.2f (%s) across the 11 rate/context metrics checked.", min(icc_preview$icc_approx), icc_preview$metric[which.min(icc_preview$icc_approx)], max(icc_preview$icc_approx), icc_preview$metric[which.max(icc_preview$icc_approx)]),
  "Most rate stats show low ICC (well under 0.2), meaning game-to-game noise dominates raw team averages -- this previews substantial shrinkage in R/06_models.R's BLUPs, especially for assisted_rate, off_tov_share, and secondchance_share.",
  "H2 note: assisted_rate's near-zero cross-sectional ICC does not contradict H2 (a trajectory claim, not a cross-sectional one) but does mean the trajectory interval on assisted rate should be reported honestly, expecting it may span zero.",
  "",
  "## 4. Outlier games",
  "",
  sprintf("%d of 364 team-games (%.1f%%) are blowouts (final margin >= 20); %d team-games are from the %d overtime games.", sum(tg$is_blowout), 100*mean(tg$is_blowout), sum(tg$is_ot), n_distinct(pbp$gameId[pbp$periodType == "OVERTIME"])),
  "Blowouts: kept, no exclusion -- pace and transition share show no meaningful shift between blowout and non-blowout team-games at the aggregate level.",
  "OT games: kept, flagged (is_ot) -- mean pace_poss is mechanically higher in OT team-games (more game time), a raw-count artifact rather than a pace signal. Spec change: R/05_features.R must carry an is_ot flag alongside pace_poss so R/06_models.R can normalize or account for it rather than treat OT team-games as unexplained outliers. Logged in PLAN.md.",
  "A handful of team-games exceed |z| > 3 on rate metrics (concentrated in transition_pts_per_poss, from small transition-possession denominators); not treated as errors, left for mixed-model shrinkage to handle rather than manual trimming.",
  "",
  "## 5. Raw trajectory eyeball plots",
  "",
  "game_index (each team's own chronological game number, not calendar date) used as the x-axis for the four AMENDMENT_01 Section 1 shortlist metrics: transition_share, transition_pts_per_poss, assisted_rate, live_ball_tov_rate.",
  sprintf("Pooled (naive OLS, not the actual trajectory model): none of the four metrics shows a statistically distinguishable linear trend through ~game 24 (all p > 0.4). Preliminary null for both H1 and H2 at the pooled level -- reported honestly per hypotheses discipline; a published null is a finding, not a failure."),
  "TOR and PDX (H2's named expansion-team cases): both show a positive (rising) raw assisted-rate slope, directionally consistent with H2. Live-ball TOV rate is directionally mixed (flat for TOR, slightly rising for PDX, the wrong direction for H2) -- the turnover side of H2 is not yet supported for either team at this raw, un-shrunk level.",
  "This is a visual/pooled gut check only. The actual test is R/06_models.R's per-team random-slope trajectory model, which a pooled fit can average away.",
  "",
  "## 6. Garbage-time decision",
  "",
  sprintf("Rule: a possession is flagged garbage_time when period >= 4 and |score margin| >= 20 at the possession's start. %d of %d possessions (%.1f%%) flagged.", sum(poss_margin$garbage_time, na.rm=TRUE), nrow(poss_margin), 100*mean(poss_margin$garbage_time, na.rm=TRUE)),
  "Decision: flag, do not exclude, in the main R/05_features.R feature table -- a 3.9% possession share is consistent with Section 4's finding of no strong aggregate distortion from blowouts, and exclusion would shrink denominators unevenly across teams. R/05_features.R should carry a garbage_time_poss_share column per team-game.",
  "R/06_models.R's AMENDMENT_01 Section 2c sensitivity check should re-run shortlist trajectory metrics excluding garbage_time possessions and confirm no finding flips direction; if one does, report it as a caveat rather than silently picking a threshold.",
  "",
  "## Spec changes forced by this EDA pass (logged in PLAN.md)",
  "",
  "1. R/05_features.R: use pace_poss (not pace_formula) as the primary pace column; keep pace_formula only as a secondary cross-check.",
  "2. R/05_features.R: add an is_ot flag alongside pace_poss so OT team-games' mechanically higher possession counts are not read as a pace signal.",
  "3. R/05_features.R: add a garbage_time_poss_share column per team-game (flag, not exclude, per the Section 6 decision).",
  "",
  "## Hypotheses registry",
  "",
  "Copied verbatim from PLAN.md. Findings in R/06_models.R and beyond are written against this registry; hypotheses are not invented after seeing results.",
  "",
  "- **H1**: Transition share and transition efficiency increase league-wide as the season progresses (reps hypothesis).",
  "- **H2**: Chemistry proxies improve with shared reps -- assisted rate rises, live-ball turnover rate falls. Expected strongest for TOR and PDX (zero shared reps at season start).",
  "- **H3**: GSV's below-average FG% reflects shot making below expectation on an acceptable shot diet, not a generation problem. Trajectory layer adds: is GSV's making trending toward expectation or flat?",
  "- **H-null discipline**: report which hypotheses the data does not support. A published null on H1 is a finding, not a failure."
)

dir.create("../output", recursive = TRUE, showWarnings = FALSE)
writeLines(notes, "../output/eda_notes.md")
message("Wrote output/eda_notes.md (", length(notes), " lines)")
```

`output/eda_notes.md` is written by the chunk above, generated directly from
the computed values in this notebook rather than transcribed by hand.
